---
title: "ST405/ST645 Bayesian Data Analysis"
subtitle: "Assignment 3: Football outcomes vs point spread"
author: "Prof. Niamh Cahill"
format: 
  html:
    embed-resources: true
    self-contained-math: true
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rjags)
library(R2jags)
library(bayesplot)
library(tidybayes)
football_dat <- read_csv("football_dat.csv")


```

## Instructions

Please submit your answers to this assignment by **November 26th, 4:00PM**. Include your JAGS model with your answers.

------------------------------------------------------------------------

### Dataset

Football experts provide a point spread for every football game as a measure of the difference in ability between the two teams. A dataset of actual outcome vs. point spread for each of 672 professional football games is provided (available on the Rstudio server). Each observation contains `spread`, x, and the outcome (favorite’s score minus underdog’s score), y. A plot of the data is shown below.

```{r, echo=TRUE}
p <- ggplot(football_dat, aes(x=spread, y=outcome)) +
  geom_point()
p
```

------------------------------------------------------------------------

### Analysis

Use JAGS to fit a linear regression model for outcome (y) as a function of point spread (x) but where the residual variance is also modeled as a function of the point spread (x) such that

$y_i|\mu,\sigma^2 \sim Normal(\mu_i, \sigma_i^2)$,

$\mu_i = \alpha + \beta x_i$,

$\sigma_i = exp(\alpha_{\sigma} + \beta_{\sigma} x_i)$.

Assume Normal(0,10\^2) priors for $\alpha$, $\beta$, $\alpha_{\sigma}$ and $\beta_{\sigma}$

This is a *heteroscedastic* linear regression, where

1.  The expected value of $y$ (outcome) for each data point $i$ is modeled with a standard linear regression describing how the mean outcome changes with the point spread.

2.  The variance of $y$ for each data point $i$ isn’t constant but changes with $x$. The log-linear formulation for $\sigma_i$ (the standard deviation of the outcome) ensures that $\sigma_i$ remains positive, which is necessary for valid variance. In the specification $\alpha_{\sigma}$ and $\beta_{\sigma}$ are parameters governing how the residual standard deviation varies with $x$.

By fitting both the mean and variance as functions of $x$ in the regression model, the model can capture two relationships:

-   The effect of the point spread on the average outcome.

-   The effect of the point spread on the variability of the outcome, potentially showing less variability in outcome values as the point spread grows.

### Questions **\[Total = 35 Marks\]**

Based on this model, complete tasks 1 - 4 below.

-   [ ] **Task 1: Model Fitting and Parameter Estimation**

 **\[8 Marks\]** Write the JAGS code and run the model to provide parameter summaries (medians and 95% credible intervals) for $\alpha$, $\beta$, $\alpha_{\sigma}$ and $\beta_{\sigma}$.

-   [ ] **Task 2: Inference**

Based on your parameter estimates,

a.  **\[2 Marks\]** does there appear to be a relationship between the outcome ($y$) and the point spread? Explain your answer.

b.  **\[3 Marks\]** does there appear to be a relationship between the residual variation ($\sigma_i$) and the point spread? Explain your answer.

-   [ ] **Task 3: Posterior Predictive Checking**

**\[4 Marks\]** Modify your JAGS code to create simulated datasets, `yrep`, used for posterior predictive checking. Then,

a.  **\[3 Marks\]**  produce a PPC plot using `ppc_stat()` to compare the distribution of minimums from `yrep` with the observed minimum from `y`;

b.   **\[4 Marks\]** produce a PPC plot using a PPC function called `ppc_intervals()`, which allows you to look at medians and uncertainty intervals for `yrep` vs a predictor variable, $x$ (use `?ppc_intervals()` to help);

c.  **\[5 Marks\]** briefly explain what you have learned from these PPC plots.

-   [ ] **Task 4: Additional Analysis**

**\[8 Marks\]** What is the median estimate and 95% prediction interval for an outcome when the point spread is 16?

------------------------------------------------------------------------
